\documentclass[11pt]{article}
\input{preamble.tex}

\title{
  Fast, scalable WCOJ graph-pattern matching on in-memory graphs in Spark\\
  \vspace{0.6cm}
  \large Master thesis draft iteration 2.0
}
\author{Per Fuchs}
\date{April 2019}

\begin{document}

\maketitle

\abstract{
  Graph pattern matching is a challenge for data processing systems like Spark because the size of the intermediary
  result required by finding the matches with standard binary join operators grow over linear with regards to the
  inputs.
  Recently popularized worst-case optimal join algorithms, WCOJ’s, seem a better match to tackle this challenge
  because they do not materialize the aforementioned large intermediary results.
  We investigate two major open questions regarding WCOJ’s in graph use cases.
  First, we develop a WCOJ specialized to graph-pattern matching, where the joins are self-joins on an edge table and compare its performance with a generic WCOJ. Second, we propose a novel method for distributed execution of such WCOJs. We show that the previously proposed methods to distribute WCOJs that would fit Spark, do not scale well to large graph patterns (five vertices and more). Based on this result, we propose to keep the edge relationship cached in compressed form on all workers but distribute the computation using a logical partitioning.
  Our approach trades off memory usage for better scalability.
  This is reasonable as most graphs today fit in main memory~\cite{snap}.
  In all, our work provides the first scalable Spark implementation of a WCOJ, very suited for graph use.
}
\newpage

\setlength{\parindent}{0em}
\setlength{\parskip}{0.7em}

\section*{Acknowledgements}
% peter for
%   super topic choice, long discussions to allow me to arrive at my dream topic
%   allow me to talk in his place at the workshop
%   being flexible with holidays for my LDR
%
The Spark integration of this work has been inspired by the IndexedDataframes~\cite{indexed-dataframes} work of Alex Uta and the supervisors
of this thesis.

\newpage
\tableofcontents
\newpage

\input{introduction.tex}
\input{background.tex}
\input{partitionings.tex}
\input{graphWCOJ.tex}
\input{implementation.tex}
\input{spark-integration.tex}
\input{experiments.tex}
\input{related-work.tex}
\input{conclusions.tex}
\clearpage

\addcontentsline{toc}{section}{References}
\printbibliography
\clearpage

\input{appendix.tex}



\end{document}
